R Python, and Ruby clients for GBIF species occurrence data

### 2017-09-06 manuscript.Rmd
added Ruby client and fixes use case problems, updated circleci badge in readme

---
layout: review, 11pt
linenumbers: true
title: "R and Python clients for GBIF species occurrence data"
author:
  - name: Scott Chamberlain
    affiliation: cstar
    email: scott(at)ropensci.org
    footnote: Corresponding author
  - name: Carl Boettiger
    affiliation: boettig
    email: carl(at)ropensci.org
address:
  - code: cstar
    address: |
      rOpenSci, Museum of Paleontology, University of California, Berkeley, CA, USA
  - code: boettig
    address: |
      rOpenSci, Department of Enivornmental Science, Policy and Management, University of California, Berkeley, CA, USA
abstract: |
      Corresponding Author:

@@

      Methods. We have developed a client to access GBIF data for each
        of the R and Python programming languages: `rgbif` and `pygbif`.

      Results. For both clients, we describe their design and utility,
        and demonstrate many use cases.

@@

There are many online services that collect and maintain specimen records.
However, Global Biodiversity Information Facility (hereafter, GBIF,
[http://www.gbif.org](http://www.gbif.org)) is the largest collection of
biodiversity records globally, currently with 643 million records, 1.6 million
taxa, 15,450 datasets from 780 publishers (as of 2016-02-09).
Many large biodiversity warehouses such as iNaturalist
([http://www.inaturalist.org](http://www.inaturalist.org)), VertNet
([http://vertnet.org](http://vertnet.org)), and USGS's Biodiversity Information
Serving Our Nation (BISON; [http://bison.usgs.ornl.gov](http://bison.usgs.ornl.gov))
all feed into GBIF.

The most important data organizational level in GBIF is the individual record. In R
using `rgbif` you'll recognize this as a row in the data.frame that is returned from
`occ_search()` or `occ_data()`. Going upstream, each record is part of a dataset,
where each dataset is submitted by an organization, organizations are organized into
nodes, datasets are published through institutions (which may be hosted at another
organization), and a network is a group of datasets (managed by GBIF).

Each record has some taxonomic name associated with it, which itself is linked to
a lot of other taxonomic data. GBIF maintains their own taxonomic data.

This organization matters because you can navigate through GBIF data through
these various organizational levels.

@@

@@

@@

`rgbif` and `pygbif` are designed following the [GBIF Application Programming Interface][gbifdocs],
or API. The GBIF API has four major components: registry, taxonomic names,
occurrences, and maps. We also include functions to interface with the OAI-PMH
GBIF service; only dataset information is available vis this service, however. We
ignore maps in both libraries as it is concerned with generating maps primarily for web
applications. Both libraries have a suite of functions dealing with each of registry, taxonomic
names, and occurrences - we'll go through each in turn describing design and example usage.

@@

With each request `rgbif` and `pygbif` make to GBIF's API, we send request headers that
tell GBIF what client the request is coming from, including what version of
the library. This helps GBIF know what proportion of requests are coming from
which client, and therefore from R vs. Python; this information is helpful for GBIF
in thinking about how people are using GBIF data.

@@

Dataset information in general is available via the OAI-PMH service,
functions in `rgbif` prefixed with `gbif_oai_`, but not available in `pygbif`
yet.

@@

Dataset functions include search, dataset metadata retrieval, and dataset metrics.
Searching for datasets is an important part of the discovery process. One can
search for datasets on the GBIF web portal. However, programmatic searching using
this package is much more powerful. Identifying datasets appropriate for a
research question is helpful as you can get metadata for each dataset, and track
down dataset specific problems, if any.

The `dataset_search()` function is one way to search for datasets. Here, we search
for the query term "oregon", which finds any datasets that have terms matching
that term.

@@

@@

Here, we search for dataset metrics for a single dataset, with uuid
`ec93a739-1681-4b04-b62f-3a687127a17f`, a checklist of the ants (Hymenoptera:
Formicidae) of the World.

@@

@@

@@

`pygbif` only has `name_backbone()` and `name_suggest()` at this time.

@@

The GBIF backbone taxonomy is used in GBIF to have a consistent way to refer
to taxonomic names throughout their services. The backbone has `r dataset_metrics("d7dddbf4-2cf0-4f39-9b2a-bb099caae36c")$distinctNamesCount` unique
names and `r dataset_metrics("d7dddbf4-2cf0-4f39-9b2a-bb099caae36c")$countByRank$SPECIES` species names. The backbone taxonomy is also a dataset with key
`d7dddbf4-2cf0-4f39-9b2a-bb099caae36c`
(http://www.gbif.org/dataset/d7dddbf4-2cf0-4f39-9b2a-bb099caae36c).

We can search the backbone taxonomy with the function `name_backbone()` in both
R and Python clients. Here, we're searching for the name _Poa_, restricting to
genera, and the family _Poaceae_, in R

@@

and in Python

@@

@@

@@

GBIF provides two ways to get occurrence data: through the
`/occurrence/search` route (see `occ_search` in `rgbif`, or
`occurrences.search` in `pygbif`), or via the
`/occurrence/download` route (many functions, see below).
`occ_search()`/`occurrences.search` is the main function for the search route,
and is more appropriate for smaller data, while
the download functions are more appropriate for larger
data requests.

GBIF imposes for any given search a limit of 200,000 records in the search
service, after which point you can't download any more records for that
search. However, you can download more records for different searches.

We think the search service is still quite useful for many people
even given the 200,000 limit. For those that need more data, we have
created a similar interface in the download functions, that should
be easy to use. Users should take note that using the download service
has a few extra steps to get data into R, but is straight-forward.

The download service, like the occurrence search service, is rate-limited.
That is, you can only have one to three downloads running simultaneously for
your user credentials. However, simply check when a download job is complete,
then you should be able to start a new download request.

@@

@@

@@

* Get a single numeric count of occurrences - `rgbif`: `occ_count()` / `pygbif`: `occurrences.count`
* Search for occurrences - `rgbif`: `occ_search()` / `pygbif`: `occurrences.count`
* A simplified and optimized version of `rgbif`: `occ_search()` or `occ_data()` / `pygbif`: `occurrences.count`
* Get occurrences by occurrence identifier - `rgbif`: `occ_get()` / `pygbif`: `occurrences.count`
* Get occurrence metadata - `rgbif`: `occ_metadata()` / `pygbif`: `occurrences.count`

@@

```{r}
(key <- name_suggest(q = 'Helianthus annuus', rank = 'species')$key[1])
occ_search(taxonKey = key, limit = 2)
```

@@

@@

@@

@@

@@

The equivalent occurrence search workhorse in `pygibf` is `occurrences.search`.

Spatial search, based on well known text format [@wkt], or a bounding box set of
four coordinates

Get only occurrences with lat/long data

@@

@@

GBIF provides optional data issues with each occurrence record. These issues fall
into many different pre-defined classes, covering issues with taxonomic names,
geographic data, and more (see `occ_issues_lookup()` to find out more information
on GBIF issues; and the same data on [GBIF's development site][gbifissues]).

`occ_issues()` provides a way to easily filter data downloaded via `occ_search()`
based on GBIF issues.

@@

There's no equivalent interface in `pygbif` yet.

@@

There's no equivalent interface in `pygbif`.

@@

# R vs. Python

@@

The following are three use cases for `rgbif`: niche modeling,
spatial change in biodiversity, and distribution mapping.

@@

The `rgbif` and `pygbif` libraries provide programmatic interfaces to GBIF's
application programming interface (API) - a powerful tool for working with
species occurrence data, and facilitating reproducible research. In fact,
the `rgbif` package has already been used in 22 scholarly publications
(as of 2015-11-14).

The `rgbif` package is relatively stable, and should not have
many breaking changes unless necessitated due to changes in the
GBIF API.

The `pygbif` library is in early development, and will
greatly benefit from any feedback and use cases.

One area of focus in the future is to attempt to solve many use
cases that have been brought up with respect to GBIF data. For
example, some specimens are included in GBIF that are located
in botanical gardens. For many research questions, researchers are
interested in "wild" type occurrences, not those in human curated
scenarios. Making removal of these occurrences easy would be
very useful, but is actually quite a hard problem. There are many
other problems like this.

@@

All scripts and data used in this paper can be found in the permanent
data archive Zenodo under the digital object identifier (DOI). This DOI
corresponds to a snapshot of the GitHub repository at
<https://github.com/sckott/msrgbif>. Software can be found at
<https://github.com/ropensci/rgbif> and <https://github.com/sckott/pygbif>,
both under MIT licenses. We thank all the users that have used `rgbif`
and `pygbif` and have given feedback and reported bugs. In addition, we
greatly appreciate the all contributions to the two packages, found
at <https://github.com/ropensci/rgbif/graphs/contributors> and
<https://github.com/sckott/pygbif/graphs/contributors>.

>>>>

---
layout: review, 11pt
linenumbers: true
title: "R Python, and Ruby clients for GBIF species occurrence data"
author:
  - name: Scott Chamberlain
    affiliation: cstar
    email: scott(at)ropensci.org
    footnote: Corresponding author
  - name: Carl Boettiger
    affiliation: boettig
    email: carl(at)ropensci.org
address:
  - code: cstar
    address: |
      rOpenSci, Museum of Paleontology, University of California, Berkeley, CA, USA
  - code: boettig
    address: |
      rOpenSci, Department of Enivornmental Science, Policy and Management, University of California, Berkeley, CA, USA
abstract: |
      Corresponding Author:

@@

      Methods. We have developed clients to access GBIF data for each
        of the R, Python, and Ruby programming languages: `rgbif`, `pygbif`,
        `gbifrb`.

      Results. For all clients we describe their design and utility,
        and demonstrate many use cases.

@@

There are many online services that collect and maintain specimen records.
However, Global Biodiversity Information Facility (hereafter, GBIF,
<http://www.gbif.org>) is the largest collection of biodiversity records
globally, currently with 820 million records, roughly 5.9 million
taxa, 36,000 datasets from 1,300 publishers (as of 2016-02-09).
Many large biodiversity warehouses such as iNaturalist
(<http://www.inaturalist.org>), VertNet (<http://vertnet.org>), and USGS's
Biodiversity Information Serving Our Nation (BISON; <http://bison.usgs.ornl.gov>)
all feed into GBIF.

The most important organizational level in GBIF occurrence data is the occurrence
record. The fields in a record vary, but include information about
taxonomy (kingdom, phylum, genus, species names) and their identifiers,
dataset metadata, and locality information including geospatial position.
Going upstream, each record is part of a dataset, where each dataset is
submitted by an organization, organizations are organized into
nodes, datasets are published through institutions (which may be hosted at another
organization), and a network is a group of datasets (managed by GBIF).

Each occurrence record has some taxonomic name associated with it, which itself is
linked to a lot of other taxonomic data - including a master taxonomic backbone
that integrates taxonomies across many taxonomic authorities.

The organization of GBIF matters because you can navigate GBIF data through
these hierarchical organizational levels - it helps to be familiar with the
terminology and how each group relates to another.

@@

Although we discuss libraries for R, Python, and Ruby here, we focus mostly on
the R library `rgbif` as it has seen the most developer and user attention,
and is the most mature.

@@

## gbifrb

`gbifrb` [@gbifrb] is a library for working with GBIF data in
the Ruby programming environment. Like Python, Ruby is a general
purpose programming language used widely in all sectors. Unlike
Python, Ruby is not used extensively in scientific disciplines.
However, a Ruby client for GBIF data can be an important tool
given how widely Ruby is used for web and web service
development.

```{ruby eval=FALSE}
gem install gbifrb
```

```{ruby eval=FALSE}
require 'gbifrb'
```

The `gbifrb` library is less mature and complete than the R and Python libraries.
It also uses an [MIT license][mit] to maximize use everywhere. `gbifrb` is developed
publicly on GitHub at <https://github.com/sckott/gbifrb>, where
development versions of the package can be installed, and bugs and feature
requests reported. Stable versions of `gbifrb` can be installed from
[Rubygems][gemgbif], the distribution network for Ruby libraries.

@@

`rgbif`, `pygbif`, and `gbifrb` are designed following the [GBIF Application Programming Interface][gbifdocs],
or API. The GBIF API has four major components: registry, taxonomic names,
occurrences, and maps. We also include functions to interface with the OAI-PMH
GBIF service; only dataset (registry) information is available via this service, however.
An interface to the GBIF maps API is in development for `rgbif`, but is non-existent for
both `pygbif` and `gbifrb`. All three libraries have a suite of functions dealing with
each of registry, taxonomic, names, and occurrences - we'll go through each in turn
describing design of the user interface and example usage.

@@

With each request `rgbif`, `pygbif`, `gbifrb` make to GBIF's API, we send request
headers that tell GBIF what library the request is coming from, including what version of
the library. This helps GBIF know what proportion of requests are coming from
which library, and therefore from R vs. Python vs. Ruby; this information is helpful
for GBIF in thinking about how people are using GBIF data.

@@

Dataset information in general is available via the OAI-PMH service,
functions in `rgbif` prefixed with `gbif_oai_`, but not available in `pygbif`
or `gbifrb` yet.

@@

Dataset functions include search, dataset metadata retrieval, and dataset metrics.
Searching for datasets is an important part of the discovery process. One can
search for datasets on the GBIF web portal. However, programmatic searching using
any of these libraries is more powerful. Identifying datasets appropriate for a
research question is helpful as you can get metadata for each dataset, and track
down dataset specific problems, if any.

The `dataset_search()` function in `rgbif` is one way to search for datasets. Here,
we search for the term "oregon", which finds any datasets that have words
matching that term.

@@

In Ruby, we can do the same. Here, search for datasets of type `OCCURRENCE`:

```{ruby eval=FALSE}
require 'gbifrb'
registry = Gbif::Registry
registry.datasets(type: "OCCURRENCE")
```

@@

Here, in R we search for dataset metrics for a single dataset, with uuid
`ec93a739-1681-4b04-b62f-3a687127a17f`, a checklist of the ants (Hymenoptera:
Formicidae) of the World.

@@

The same in Ruby:

```{ruby eval=FALSE}
require 'gbifrb'
registry = Gbif::Registry
registry.dataset_metrics(uuid: 'ec93a739-1681-4b04-b62f-3a687127a17f')
```

@@

And in Ruby:

```{ruby}
require 'gbifrb'
registry = Gbif::Registry
registry.networks(limit: 10)
```

@@

`pygbif` and `gbifrb` have all the same functions, except the name parser goes by
`name_parser()` in `pygbif` and `gbifrb`.

@@

The GBIF backbone taxonomy is used in GBIF to have a consistent way to refer
to taxonomic names throughout their services. The backbone has `r dataset_metrics("d7dddbf4-2cf0-4f39-9b2a-bb099caae36c")$distinctNamesCount` unique
names and `r dataset_metrics("d7dddbf4-2cf0-4f39-9b2a-bb099caae36c")$countByRank$SPECIES`
species names. The backbone taxonomy is also a dataset with key
`d7dddbf4-2cf0-4f39-9b2a-bb099caae36c`
(<https://www.gbif.org/dataset/d7dddbf4-2cf0-4f39-9b2a-bb099caae36c>).

We can search the backbone taxonomy with the function `name_backbone()` in all
thee clients. Here, we're searching for the name _Poa_, restricting to
genera, and the family _Poaceae_, in R

@@

And in Python

@@

And in Ruby

```{ruby eval=FALSE}
require 'gbifrb'
species = Gbif::Species
res = species.name_backbone(name: 'Poa', rank: 'genus', family: 'Poaceae')
res.select { |k,v| k.match(/usageKey|kingdom/) }
```

@@

And in Ruby

```{ruby eval=FALSE}
require 'gbifrb'
species = Gbif::Species
species.name_suggest(q: 'Pum', limit: 6)
```

@@

GBIF provides two ways to get occurrence data: through the
`/occurrence/search` route (see `occ_search` in `rgbif`,
`occurrences.search` in `pygbif`, `Occurrences.search` in gbifrb), or via the
`/occurrence/download` route (many functions, see below).

`occ_search()`/`occurrences.search`/`Occurrences.search` are the main functions
for the search route, and are more appropriate when you want less data, while
the download functions are more appropriate for larger data requests.

Small vs. large amounts of data of course is all relative. GBIF imposes for any
given search a limit of 200,000 records in the search
service, after which point you can't download any more records for that
search. However, you can download more records for different searches.

We think the search service is still quite useful for many people
even given the 200,000 limit. For those that need more data, we have
created a similar interface in the download functions that should
be easy to use with minimal work. Users should take note that using the
download service has a few extra steps to get data into R, but is
straight-forward.

The download service, like the occurrence search service, is rate-limited.
That is, you can only have one to three downloads running simultaneously for
your user credentials. However, simply check when a download job is complete,
then you can start a new download request. See "Queueing Download Requests"
below for help automating many download requests in R.

@@

The Ruby library `gbifrb` does not yet have occurrene download functionality.

@@

#### Queueing Download Requests

FILL IN TEXT HERE ...

@@

* Get a single numeric count of occurrences - `rgbif`: `occ_count()` / `pygbif`: `occurrences.count` / `gbifrb`: `Occurrences.count`
* Search for occurrences - `rgbif`: `occ_search()` / `pygbif`: `occurrences.search` / `gbifrb`: `Occurrences.search`
* A simplified and optimized version of `rgbif`: `occ_search()` or `occ_data()` / none / none
* Get occurrences by occurrence identifier - `rgbif`: `occ_get()` / `pygbif`: `occurrences.get` / `gbifrb`: `Occurrences.get`
* Get occurrence metadata - `rgbif`: `occ_metadata()` / `pygbif`: various / `gbifrb`: various

@@

```{r}
library(rgbif)
(key <- name_suggest(q = 'Helianthus annuus', rank = 'species')$key[1])
occ_search(taxonKey = key, limit = 2)
```

__Ruby__

```{ruby eval=FALSE}
require 'gbifrb'
species = Gbif::Species
occ = Gbif::Occurrences
key = species.name_suggest(q: 'Helianthus annuus', rank: 'species')['data'][0]['key']
occ.search(taxonKey: key, limit: 2)
```

@@

__Ruby__

```{ruby eval=FALSE}
occ.search(scientificName: 'Ursus americanus')
```

@@

__Ruby__

```{ruby eval=FALSE}
species = Gbif::Species
occ = Gbif::Occurrences
splist = ['Cyanocitta stelleri', 'Junco hyemalis', 'Aix sponsa']
keys = [ species.name_suggest(x)['data'][0]['key'] for x in splist ]
occ.search(taxonKey: keys, limit: 5)
```

@@

__Ruby__

```{ruby eval=FALSE}
occ = Gbif::Occurrences
# well known text
occ.search(geometry: 'POLYGON((30.1 10.1, 10 20, 20 40, 40 40, 30.1 10.1))', limit: 20)
# bounding box
occ.search(geometry: '-125.0,38.4,-121.8,40.9', limit: 20)
```

@@

__Ruby__

```{ruby eval=FALSE}
occ = Gbif::Occurrences
occ.search(hasCoordinate: true, limit: 5)
```

@@

@@

__Ruby__

```{ruby eval=FALSE}
occ = Gbif::Occurrences
occ.search(hasGeospatialIssue: true, limit: 5)
```

@@

GBIF provides optional data issues with each occurrence record. These issues fall
into many different pre-defined classes, covering issues with taxonomic names,
geographic data, and more (see `rgbif::occ_issues_lookup()` to find out more information
on GBIF issues; and the same data on [GBIF's development site][gbifissues]).

`rgbif::occ_issues()` provides a way to easily filter data downloaded via `rgbif::occ_search()`
based on GBIF issues.

@@

There's no equivalent interface in `pygbif` or `gbifrb` yet.

@@

There's no equivalent interface in `pygbif` or `gbifrb`.

@@

# R vs. Python vs. Ruby

@@

The following are three use cases for the R library `rgbif`: niche modeling,
spatial change in biodiversity, and distribution mapping.

@@

The `rgbif`, `pygbif`, and `gbibrb` libraries provide programmatic interfaces
to GBIF's application programming interface (API) - a powerful tool for
working with species occurrence data, and facilitating reproducible research.
In fact, the `rgbif` package has already been used in more than 20 scholarly
publications (as of 2016-08-10), including [@Amano_2016, @Bartomeus_2013, @Barve_2014, @Bone_2015, @Collins_2015, @Drozd_2013, @Kong_2015, @Richardson_2015, @Turner_2015, @Verheijen_2015, @Zizka_2015, @Butterfield_2016, @Dellinger_2015, @Feitosa_2015, @Malhado_2015, @Werner_2015, @Robertson_2016, @Davison_2015, @Janssens_2016].

The `rgbif` package is relatively stable, and should not have
many breaking changes unless necessitated due to changes in the
GBIF API. However, it will gain function(s) to work with the maps
API in the near future.

The `pygbif` and `gbifrb` libraries are in early development, and will
greatly benefit from any feedback and use cases.

One area of focus in the future is to attempt to solve many use
cases that have been brought up with respect to GBIF data. For
example, some specimens are included in GBIF that are located
in botanical gardens. For many research questions, researchers are
interested in "wild" type occurrences, not those in human curated
scenarios. Making removal of these occurrences easy would be
very useful, but is actually quite a hard problem. There are many
other problems like this, for which these three libraries will help
in making more efficient and reproducible.

@@

All scripts and data used in this paper can be found in the permanent
data archive Zenodo under the digital object identifier (DOI). This DOI
corresponds to a snapshot of the GitHub repository at
<https://github.com/sckott/gbifms>. Software can be found at
<https://github.com/ropensci/rgbif>, <https://github.com/sckott/pygbif>,
and <https://github.com/sckott/gibfrb>, all under MIT licenses. We thank
all the users that have used `rgbif`, `pygbif`, and `gbifrb` and have
given feedback and reported bugs. In addition, we greatly appreciate
all the contributors to the three libraries, found at
<https://github.com/ropensci/rgbif/graphs/contributors>,
<https://github.com/sckott/pygbif/graphs/contributors>, and
<https://github.com/sckott/gbifrb/graphs/contributors>.
